### 2025-04-10 docs/thesis_proposal.tex
syncing...

\documentclass[paper=a4, fontsize=11pt,twoside]{scrartcl}       % KOMA

\usepackage[a4paper,pdftex]{geometry}   % A4paper margins
\setlength{\oddsidemargin}{5mm}                 % Remove 'twosided' indentation
\setlength{\evensidemargin}{5mm}

\usepackage[english]{babel}
\usepackage[font=footnotesize, labelfont=it]{caption}
\usepackage[protrusion=true,expansion=true]{microtype}
\usepackage{mathtools}
\usepackage{graphicx}
\usepackage{wrapfig}
\usepackage{subfig}
\usepackage{amsfonts}
\usepackage[autocite = superscript, backref = true, sorting = none]{biblatex}
\usepackage{csquotes}
\addbibresource{proposal_bib.bib}
\usepackage{hyperref}
\hypersetup{
        colorlinks=true,
        linkcolor=blue,
        filecolor=magenta,
        urlcolor=blue,
        citecolor=blue
        }
\urlstyle{same}

\setcounter{secnumdepth}{0} % no section numbering
\pagenumbering{gobble} % no page numbering

\begin{document}

    \section{Background}

    Fitness landscape research typically models adaptation on static landscapes, where the fitness values and dimensionality of genotype space are fixed. Biologically, this simulates a population adapting to a new, constant environment, stopping once a fitness peak is reached. However, this approach does not incorporate other phenomena that occur during evolution, such as changes in the environment (otherwise known as "seascapes"), or particularly, changes in the organism's genome size (genome evolution). Since, over long time scales, both adaptation and genome evolution are processes relevant to the evolution of the organism, it is important to create a model that can incorporate both to augument our understanding of how genomes expand and streamline, and ultimately, how new genes and genomic complexity might arise. Furthermore, simulation of such an adaptive system gives rise to path dependency (a high-order autoregressive/history-dependent markov process), for which diverse literature across ecology \cite{fukamiHistoricalContingencyCommunity2015}, evolution \cite{kauffmanProlegomenonPatternsEvolution2014}, public policy \cite{petersPoliticsPathDependency2005}, and strategic management \cite{siggelkowFirmsSystemsInterdependent2011} has noted difficulty in creating general predictive frameworks. I will utilize an extension of the Rough Mount Fuji model combined with a Counter-Based Random Number Generator to model population dynamics on these "Expanding Landscapes" and explore the following questions:

    \section{Research Questions}

	\subsection{Simulation/Experimental}

    	    \begin{itemize}

		    \item What balance is reached between adaptation and genome evolution? Is one mechanism of navigating the landscape preferred to the other? How can this relationship be quantified?
		    \item How does this balance shift as the ruggedness of the landscape changes? As the strength of drift changes?
		    \item What diversity do we see arise in the population (in terms of hamming distance between genotypes, and number of unique genomes)? Is this generated continuously, or is there burstiness? \cite{gohBurstinessMemoryComplex2008}
		    \item How repeatably are certain maxima reached, or certain loci unlocked? Under which conditions, and to what degree, does path dependency determine the order of events observed?

	    \end{itemize}

	\subsection{Theoretical}

	    \begin{itemize}

		    \item How does the formulation of the model fit into a framework of path dependent processes from literature? \cite{jacksonModelingMeasuringDistinguishing2012}
		    \item What does a historical contingency look like in this model?
		    \item Is there an interpretation of the model that is compatible with experimental evidence of gene evolution? That is, such that new sequences are selectively near-neutral and generally fix through drift rather than selection. \cite{lynchFrailtyAdaptiveHypotheses2007}
		    \item Can we extend this to ecological community models? Technically all genomes and genotypes are genetically compatible. Can we modify the model to restrict reproduction between only genotypes a maximum hamming distance, and if so, do we see stable equilibria of multiple populations emerge? Additionally, niche theory predicts that the order of community assembly matters. Would the order of loci unlocking then also dictate the community composition?

	    \end{itemize}

    \section{Timeline}

        \begin{description}

		\item [February 17] Thesis start - background research \& drafting model 
		\item [March 7] Master thesis start form due to program admin
		\item [March 28] THEE Project proposal due to Claudia
		\item [March 31] Deadline to implement preliminary version of RMF model
		\item [April 1] Reduce hours to 60\% to work on Suman's project (until August)
		\item [May 5] Goal date to formalize theoretical basis 
		\item [May 5] Begin drafting thesis outline (Introduction, Materials \& Methods)
		\item [August 1] Return to 100\% work
		\item [Sometime in August] (Mostly) stop research, 100\% writing
		\item [August 14] \textbf{FS25 Deadline} Submit thesis to Claudia \& Stephan for grading
		\item [September 14] \textbf{FS25 Deadline} Final thesis submission to Dean's office
		\item [September 17] \textbf{HS25 Deadline} Submit thesis to Claudia \& Stephan for grading
		\item [October 17] \textbf{HS25 Deadline} Final thesis submission to Dean's office

	\end{description}

	If I submit by the September deadline and not in October, I avoid needing to register for another semester. I am not sure if it is feasible to get the writing done in 2 weeks beginning of August, and/or if I will have the right experimental results at that time.

	\printbibliography

\end{document}

### 2025-08-28 docs/thesis_manuscript/manuscript.tex
now with custom starting genotype/genome!

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% LaTeX Template: Project Titlepage
%
% Source: http://www.howtotex.com
% Date: April 2011
%
% This is a title page template which be used for articles & reports.
%
% Feel free to distribute this example, but please keep the referral
% to howtotex.com
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% --------------------------------------------------------------------
% Preamble
% --------------------------------------------------------------------

\documentclass[paper=a4, fontsize=11pt,twoside]{scrartcl}       % KOMA

\usepackage[a4paper,pdftex]{geometry}   % A4paper margins
\setlength{\oddsidemargin}{5mm}                 % Remove 'twosided' indentation
\setlength{\evensidemargin}{5mm}

\usepackage[english]{babel}
\usepackage[font=footnotesize, labelfont=it]{caption}
\usepackage[protrusion=true,expansion=true]{microtype}
\usepackage{mathtools}
\usepackage{graphicx}
\usepackage{wrapfig}
\usepackage{subfig}
\usepackage{amsfonts}
\usepackage[backref=true, sorting=none]{biblatex}
\usepackage{csquotes}
\addbibresource{bibliography.bib}
\usepackage{hyperref}
\hypersetup{
        colorlinks=true,
        linkcolor=blue,
        filecolor=magenta,
        urlcolor=blue,
        citecolor=blue
        }
\urlstyle{same}

% --------------------------------------------------------------------
% Definitions (do not change this)
% --------------------------------------------------------------------

\newcommand{\HRule}[1]{\rule{\linewidth}{#1}}   % Horizontal rule

\makeatletter                                                   % Title
\def\printtitle{%
    {\centering \@title\par}}
\makeatother

\makeatletter                                                   % Author
\def\printauthor{%
    {\centering \large \@author}}
\makeatother

% --------------------------------------------------------------------
% Metadata (Change this)
% --------------------------------------------------------------------

\title{
                        \HRule{0.5pt} \\                                                % Upper rule
                        \LARGE \textbf{\uppercase{A Computationally Tractable Model for the Evolution of a Genome's Size and its Growing Fitness Landscape}}      % Title
                        \HRule{2pt} \\ [0.5cm]          % Lower rule + 0.5cm spacing
                        \normalsize \today                      % Todays date
                }

\author{
                \textbf{Master's Thesis}\\
                Faculty of Science, University of Bern\\
                \vspace{1cm}
                Handed in by \\
                \textbf{Jacob Hugh Riina} \\
                \vfill
                Supervisors\\
                Prof. Dr. Claudia Bank\\
                Dr. Stephan Peischl\\
}

\begin{document}

% ------------------------------------------------------------------------------
% Maketitle
% ------------------------------------------------------------------------------

\thispagestyle{empty}           % Remove page numbering on this page

\printtitle                                     % Print the title data as defined above
        \vfill
\printauthor                            % Print the author data as defined above
\newpage

% ------------------------------------------------------------------------------
% Begin document
% ------------------------------------------------------------------------------

\setcounter{page}{1}            % Set page numbering to begin on this page

\section*{Introduction}

\subsection*{Genome Evolution}

Since life's origins approximately 3.5 billion years ago, it has generally increased in complexity, and understanding the processes that generate this complexity is one of the main goals of evolutionary biology. One such mechanism of generating complexity is through genome evolution: by changing the size of their genomes, the earliest living cells - those with a cell membrane, which encode and pass on information through their DNA - were able to create \textit{evolutionary innovations}: new metabolism, physiology or regulation, which augumented their ability to thrive in a variety of environments. At the increased cost of replicating and proofreading more DNA, organisms have since been able to evolve new genes allowing them to metabolize different carbon, nitrogen, or complex nutrient sources, various types of receptors and secretion systems to interact, competitively or mutualistically, with others, and the internal regulatory networks required for them to decide which behaviors are optimal under which circumstances, or even to coordinate multicellularity. Several billion years later, we see the consequences of this incessant evolutionary process: the stunning diversity amongst organisms in the tree of life, each a product of the series of environments they have adapted to. We can say, then, that the content of an organism's genome is not solely the result of how it has evolved to fit its current environment, but also the entire history of environments that its ancestors have evolved in; as always, "Nothing in biology makes sense except in the light of evolution".\\

Understanding this facet of evolution is [difficult]. Unlike with [allelic mutation], genome evolution occurs up to several orders of magnitude less freqently, and is more heterogenous in its manifestation. We define here genome evolution as a mutational process resulting in either the expansion or reduction in the size of the [functional/coding] genome, stemming from one of several causes: amongst genome-increasing mechanisms, there exist duplications from repeat region replication errors, transposon activity, or whole genome duplications, as well as genes arising from noncoding sequence through de novo gene birth, while genome-reducing mechanisms are pseudogenization and deletions via various replication errors. here are some more specifics about the expansion/reduction process. here is why they are difficult to observe or create experiments to test. here is why is would be beneficial to create models that allow us to ask [] questions that bridge the gap between theory and experiment. 
Owing to a large protein fold space, evolution has generally favored mechanisms of duplication and neofunctionalization amongst new genes through horizontal gene transfer, transposon activity, or whole-genome duplications. (by doing so the cell acheives two things: 1. relieve pleiotropic effects, 2. lessen time to navigate sequence space). Alternatively, unused regions of the geome may also see genes spontaneously arise. (this process is much slower and harder to observe). In either case, these new genes are expected to accumulate neutrally, and over time have the opportunity to gain beneficial functionality. 
Although organisms are effective at removing unnecessary genes through pseudogenization, (this is a slow process, and still does not account for historical contingencies)
Different organisms may happen to adapt to the same environment in different way

We do however see in even shorter evolution experiments that not only mutation of alleles, but also genomic abberations appear that explain increases in fitness. (noted rate of nucleotide gain per generation stats here)
\subsection*{Fitness Landscape Theory}

As has been noted by studies of the \textit{de novo} evolution of genes, their emergence out of unused genetic substrate remains difficult to observe experimentally. (why? duplication vs de novo navigation of landscapes). There is thus a usefulness to generation of models that allow us to form hypotheses about these difficult to test systems.

Although fitness landscape theory has been used extensively to model adaptation on static genomes, it is less well developed on the scale of long term evolution, where changes in the genome size also arise. With this work we extend the Rough Mount Fuji model to incorporate genome evolution alongside allelic mutation

\section*{Methods}

\subsection*{The Rough Mount Fuji Model}

The Rough Mount Fuji (RMF) model allows for tunable epistasis \cite{neidhartAdaptationTunablyRugged2014} and so enables the investigation its effects as landscapes scale. In this model we represent genotypes as binary vectors, and their permutations in genotype space construct a network where each has a hamming distance of 1 from its neighbors, and the degree of each node is equivalent to the number of loci $L$ in the model (see Figure~\ref{fig:genotype_space}).

The assignment of fitness values to genotypes entails sampling two probabilistically determined components: Locus-wise additive effects $a = (a_1, \cdots, a_i) \sim \text{iid. } \mathcal{N}(0, 1)$ are drawn from a normal distribution, and each genotype $g$ recieves a unique epistatic component $e_g \sim \mathcal{N}(0, \sigma_e)$, also drawn from a normal distribution.  Fitness values for some arbitrary genotype can then be assigned as the sum of the product of the additive effects and the genotype, plus the epistatic component, as an exponential: $$f_g = exp{(e_g + \sum_{i=0}^{L} a_i * g_i)}$$ Epistasis is represented in the ratio between the variance of additive components and epistatic components, and we control epistasis on the landscape by varying the epistatic $\sigma_e$ - in cases where the epistatic component is small the additive component dominates and the landscape becomes easily predictable, whereas when epistatic effects are large and outweigh the additive component, there is virtually no correlation between different genotypes and ruggedness is high. Now able to reproducibly retrieve some epistatic component for any genotype by providing the CBRNG with the base-10 representation of that binary genotype, we are no longer obligated to generate fitnesses for all genotypes on the landscape in advance and instead need only do so upon encountering that genotype during simulation.

A discrete-generation RMF model with Wright-Fisher dynamics was implemented based on \cite{liRapidAdaptationRecombining2024} and previous software \cite{amadoSTUNForwardtimeSimulation2023}. We model a static $N$ total individuals mutating and reproducing simultaneously in each generation, with genotype frequencies governed by binomial and multinomial sampling, respectively. To initialize the simulation, we specify the number of loci in the model and consequently the length of the vector of additive effects to generate, the number of randomly distributed initial genotypes on the landscape and the number of individuals per genotype, the simulation length, the variance $\sigma$ of the epistatic component, and the mutation rate $\mu$ - the probability for an individual to generate a mutant of hamming distance 1 from its parent. An initial report table is then generated with a description of initial genotypes on the landscape, their population sizes, and fitnesses. We proceed to instantiate subsequent generations by first generating mutants from each extant genotype with population $\mathbb{P}$ through binomial sampling: $\text{Binomial}(n = \mathbb{P}, p = \mu)$. We adjust the corresponding population counts and calculate fitnesses for new genotypes if necessary, and then assign from $i$ extant genotypes for all individuals in the new generation of size $N$ new genotypes according to relative fitness: $\text{Multinomial}(N, p^{\prime} = (p^{\prime}_1, \cdots, p^{\prime}_i))$ given $p^{\prime}_i = p_i * \frac{\omega_i}{\bar{\omega}}$, where $p_i$ is some genotype's current frequency, $\omega_i$ is its fitness, and $\bar{\omega}$ is the average fitness across all extant genotypes. Therefore, highly fit mutants gain an advantage during sampling and are less likely to be eliminated through drift. The simulation is run for the specified number of generations and upon terminating a final report is also provided (Figure~\ref{fig:report}).

\subsection*{Counter-Based Random Number Generation}

The facile generation of random epistatic components was performed using a Julia implementation \cite{JuliaRandomRandom123jl2024} of the Random123 package from D.E. Shaw Research \cite{salmonParallelRandomNumbers2011}. Particularly, the ARS1x generator was used, with a period of $2^{128}$, this being the number of loci that can be modeled in a genotype network without reseeding.

\subsection*{Scripting}

\section*{Results}

\subsection*{Additive Limit}

\subsection*{Uncorrelated Limit}

\subsection*{Time-Dynamic Outputs}

\subsection*{Endpoint Outputs}

Does endpoint exist under this model?

\section*{Discussion}

\section*{Code and Data Availability}

\printbibliography

\end{document}

### 2025-08-28 code/Utils.jl

@@ -196,7 +196,7 @@ end
 """
 function to simulate a wright-fisher population with non-overlapping generations on a rough mount fuji landscape. Returns a dataframe with columns -> [:Step, :Genotype, :Genome, :Population]. Note that when printing the outputs from dataframes, UInt128s are encoded as such, but displayed in base 10. 
 """
-function simulate(loci, init_active_loci, max_init_genotype_bits, total_population, epistasis, μ, M, simulation_length, rng_init_genome, rng_init_genotype, rng_default, rng_mutation, additive_effects; model = "rmf")
+function simulate(loci, init_active_loci, max_init_genotype_bits, total_population, epistasis, μ, M, simulation_length, rng_init_genome, rng_init_genotype, rng_default, rng_mutation, additive_effects; model = "rmf", init = nothing)
     # Initialize some parameters and objects
     num_genotypes = 1 # leaving this in for posterity
     genotype_dictionary = Dict{Tuple, Genotype}() # keys are stored as a tuple of the genotype(integer) and its genome, and the genotype object as values
@@ -205,6 +205,9 @@ function simulate(loci, init_active_loci, max_init_genotype_bits, total_populati
     bits_to_flip = sample(rng_init_genome, 1:loci, init_active_loci, replace = false) # generate a vector of unique loci to unlock at start, limited to within l total loci
     for locus in bits_to_flip
         init_genome = flip_bit(init_genome, locus) # flip them, while retaining UInt128 - this is important as to not lose any information about the latent genome space!
+    end
+    if !isnothing(init) # If you've passed in some specific genotype/genome pair list, utilize this instead
+        init_genome = init[2]
     end
     #println("Initial Genome: ", init_genome, " \n\t with active bits:", bits_to_flip)
 
@@ -216,6 +219,9 @@ function simulate(loci, init_active_loci, max_init_genotype_bits, total_populati
         init_genotype::UInt128 = 0 # assume it exists as a 128bit, since it's the max size and it gets cast to this in the cbrng anyway
         for locus in sample(rng_init_genotype, bits_to_flip[1:max_init_genotype_bits], rand(rng_init_genotype, 0:max_init_genotype_bits), replace = false)  # generate a *random number* of bits that are active in the initial genome to be flipped
             init_genotype = flip_bit(init_genotype, locus)
[…]

### 2025-08-29 docs/thesis_manuscript/manuscript.tex
including the dynamics analysis

@@
 Unlike with [allelic mutation], genome evolution occurs up to several orders of magnitude less freqently, and is more heterogenous in its manifestation.
-We define here genome evolution as a mutational process resulting in either the expansion or reduction in the size of the [functional/coding] genome, stemming from one of several causes: amongst genome-increasing mechanisms, there exist duplications from repeat region replication errors, transposon activity, or whole genome duplications, as well as genes arising from noncoding sequence through de novo gene birth, while genome-reducing mechanisms are pseudogenization and deletions via various replication errors.
-here are some more specifics about the expansion/reduction process.
-here is why they are difficult to observe or create experiments to test.
-here is why is would be beneficial to create models that allow us to ask [] questions that bridge the gap between theory and experiment.
+We define here genome evolution as a mutational process resulting in either the expansion or reduction in the size of the [functional/coding] genome, arising from one of several causes: amongst genome-increasing mechanisms, there exist duplications from repeat region replication errors, transposon activity, or whole genome duplications, as well as genes arising from noncoding sequence through de novo gene birth, while genome-reducing mechanisms include pseudogenization and deletions via various types of replication errors.
+Evolution has been observed to utilize both genome reduction and expansion spur adaptation to increase the organism's \textit{fitness}, conceptualized as some metric representing success e.g.
+growth rate within an environment.
+However, some of the more impactful processes such as neofunctionalization of duplications, de novo gene birth, and WGD, remain difficult to experimentally induce and test.
+(why?) There is thus a usefulness to the creation of mathematical models that bridge the gap between theory and experiment and allow us to form hypotheses about these phenomena and their roles in long-term evolution.\\
+
+Given a model of genome size evolution, what are we interested in testing? (path dependency)\\
+
+here are some more specifics about the expansion/reduction process. here is why they are difficult to observe or create experiments to test.
 Owing to a large protein fold space, evolution has generally favored mechanisms of duplication and neofunctionalization amongst new genes through horizontal gene transfer, transposon activity, or whole-genome duplications.
@@
 duplication vs de novo navigation of landscapes).
-There is thus a usefulness to generation of models that allow us to form hypotheses about these difficult to test systems.
 

### 2025-08-29 code/Utils.jl

@@ -293,7 +293,7 @@ Function to process the dataframe output of simulations to obtain relevant time-
 4. The average hamming distance between all genotypes per genome per generation -> [:Step, :AvgGenotypeRichness]
 5. The weighted average fitness of all individuals in a generation [:Step, :AverageFitness]
 6. The weighted average genome size of all individuals in a generation [:Step, :AverageGenomeSize]
-7. The generations in which selective sweeps occur, defined as when a new genotype OR genome/genotype combination first occupies more than 50% of the population -> [:Step, :Genotype, :Genome, :Pop, :Genotype, :GenotypeChange, :GenomeChange]
+7. The generations in which selective sweeps occur, defined as when a new genotype OR genome/genotype combination first occupies more than 50% of the population -> [:Step, :Genotype, :Genome, :Pop, :Genotype, :GenotypeChange, :GenomeChange, :Fitness, :GenomeSize]
 """
 function process_data(data::DataFrame, μ, M, additive_effects, σ_epi)
     
@@ -337,6 +337,7 @@ function process_data(data::DataFrame, μ, M, additive_effects, σ_epi)
         Ref(σ_epi),
         Ref("rmf")
     )
+    sweeps[!, :CurrentGenomeSize] = sum.(digits.(sweeps.Genome, base = 2))
 
     return (df_counts, df_full_steps,
             df_genome_counts, df_per_genome_genotypes,

### 2025-08-29 code/intermediate_analysis.R

@@ -1,0 +1,99 @@
+library(tidyverse)
+library(ggridges)
+library(ggpmisc)
+setwd("~/mastersarbeit/thesis/code")
+
+# Function for the solution from Suman that yields the expectation for the distribution
+# of fitness peaks on a landscape
+expected_max_fitness <- function(L, add, epi){
+  return(exp((2 * L * (((L / 2) * add^2) + epi^2) * log(2))^0.5))
+}
+
+################################################################################
+# Begin with the static landscape
+################################################################################
+
+filepath_df01 <- "outputs/data/"
+df01 <- read.csv(filepath_df01) %>% 
+  mutate("percent_max" = Final_fitness * 100 / expected_max_fitness(Loci, 0, 1))
+
+################################################################################
+# Now the expanding landscape
+################################################################################
+
+filepath_df02 <- "outputs/data/09_intermediate_expanding_landscape_20250827_2339.csv"
+df02 <- read.csv(filepath_df02) %>% 
+  mutate("percent_max" = Final_fitness * 100 / expected_max_fitness(Final_genome_size, 0.1, 0.1),
+         "Final_genome_size" = round(Final_genome_size, 0))
+
+# On a max 20 loci landscape, we started replicates at 5, 6, 7, 8, 9, & 10 active loci.
+# What's the distribution of final sizes?
+
+ggplot(df02, aes(x = Final_genome_size)) +
+  geom_histogram(binwidth = 1)
+
+# 7, 8, 9 are all good candidates. 10 will let us analyze the nonchanging genomes as well.
+
+ggplot(filter(df02, Final_genome_size == 6), aes(x = as.factor(Loci), y = Final_fitness, group = as.factor(Loci))) +
+  geom_boxplot()
+
+################################################################################
+# Looking at dynamics (through the series of selective sweeps) of expanding landscapes
+################################################################################
+
+filepath_df03 <- "outputs/data/10_dynamics_intermediate_expanding_landscape_20250828_1702.csv"
+df03 <- read.csv(filepath_df03) %>% 
+  mutate("FinalGenomeSize" = round(FinalGenomeSize, 0),
[…]

### 2025-09-01 docs/thesis_manuscript/manuscript.tex
generate plots now generates better resolution plots

--- 2025-08-28 docs/thesis_manuscript/manuscript.tex
@@
 
-Understanding this facet of evolution is [difficult].
-Unlike with [allelic mutation], genome evolution occurs up to several orders of magnitude less freqently, and is more heterogenous in its manifestation.
-We define here genome evolution as a mutational process resulting in either the expansion or reduction in the size of the [functional/coding] genome, stemming from one of several causes: amongst genome-increasing mechanisms, there exist duplications from repeat region replication errors, transposon activity, or whole genome duplications, as well as genes arising from noncoding sequence through de novo gene birth, while genome-reducing mechanisms are pseudogenization and deletions via various replication errors.
-here are some more specifics about the expansion/reduction process.
-here is why they are difficult to observe or create experiments to test.
-here is why is would be beneficial to create models that allow us to ask [] questions that bridge the gap between theory and experiment.
-Owing to a large protein fold space, evolution has generally favored mechanisms of duplication and neofunctionalization amongst new genes through horizontal gene transfer, transposon activity, or whole-genome duplications.
-(by doing so the cell acheives two things: 1.
-relieve pleiotropic effects, 2.
-lessen time to navigate sequence space).
-Alternatively, unused regions of the geome may also see genes spontaneously arise.
-(this process is much slower and harder to observe).
-In either case, these new genes are expected to accumulate neutrally, and over time have the opportunity to gain beneficial functionality.
-Although organisms are effective at removing unnecessary genes through pseudogenization, (this is a slow process, and still does not account for historical contingencies)
-Different organisms may happen to adapt to the same environment in different way
+Understanding this facet of evolution is difficult.
+Unlike with the point mutations that are typically observed in every generation, genome evolution occurs up to several orders of magnitude less frequently, and is more heterogenous in its manifestation.
+We define here genome evolution as a mutational process resulting in either the expansion or reduction in the size of the functional genome, arising from one of several causes: amongst genome-increasing mechanisms, there exist duplications from repeat region replication errors, transposon activity, or whole genome duplications (WGD), as well as genes arising from noncoding sequence through de novo gene birth, while genome-reducing mechanisms include pseudogenization and deletions via various types of replication errors.
+Evolution has been observed to utilize both genome reduction and expansion in order to spur adaptation to increase the organism's \textit{fitness}, conceptualized as some metric representing success e.g.
+growth rate within an environment.
+However, more impactful processes such as neofunctionalization of duplications, de novo gene birth, and WGD, remain slow to induce and test, often in experiments running for years or decades.
+There is thus a usefulness to the creation of mathematical models that bridge the gap between theory and experiment and allow us to form hypotheses about these phenomena and their impact over long evolutionary timescales.\\
 
-We do however see in even shorter evolution experiments that not only mutation of alleles, but also genomic abberations appear that explain increases in fitness.
-(noted rate of nucleotide gain per generation stats here)
+One such [question] naturally arising from the study of this type of adaptive system is regarding the effects of \textit{path dependency} during genome evolution: how does the order and timing in which evolutionary innovations arise affect evolutionary possibilities far in the future, potentially in different environments?
+For instance, the nitrogenase metalloenzyme - the main driver of bioavailable nitrogen cycling, which utilizes multiple oxygen-labile iron-sulfur clusters - emerged under the earth's early atmospheric conditions and favored adaptations suited to a reducing atmosphere.
+Faced with a later change to an oxidizing and highly destructive environment, organisms with this metabolic innovation, incapable of decommitting from fixing their own nitrogen to sustain cell growth, instead evolved a set of highly complicated cellular pathways to continue catalysis under only anoxic conditions, and protect the enzyme's metallic core from molecular oxygen.
+No other method of creating bioavailable nitrogen in an oxygenic cellular environment has ever been discovered.
+What if nitrogen fixation had first evolved after the oxygenation of the atmosphere?
+Is it even possible, and if so, how would the enzyme differ from today's?
+Evolution appears to be rife with examples of historical contingencies, spanning from within microbial metabolism up to macroscale physiology[cite giraffe].
+[these are questions about the evolution of genomic context over time, and therefore a good starting point is from mathematical models that allow for the modeling of context, and how it can evolve.
+We thus turn to fitness landscapes.] how do different cellular contexts create different evolutionary outcomes, which then lead to even more different contexts, which lead to even different outcomes?
+
 \subsection*{Fitness Landscape Theory}
 
-As has been noted by studies of the \textit{de novo} evolution of genes, their emergence out of unused genetic substrate remains difficult to observe experimentally.
-(why?
-duplication vs de novo navigation of landscapes).
-There is thus a usefulness to generation of models that allow us to form hypotheses about these difficult to test systems.
-
-Although fitness landscape theory has been used extensively to model adaptation on static genomes, it is less well developed on the scale of long term evolution, where changes in the genome size also arise.
-With this work we extend the Rough Mount Fuji model to incorporate genome evolution alongside allelic mutation
+The fitness landscape was first introduced in 19xx by Sewall Wright to illustrate xyz.
+With each unique combination of decisions there can be assigned a fitness value, and so we can generate a topology from a network of possible decision combinations.
+Fitnesses for contexts on the landscape may be higher or lower than expected from the sum of their individual components, leading to "ruggedness", otherwise known as \textit{epistasis}: nonadditivity resulting from higher-order interactions between components, that ultimately leads to, in the case of evolution, adapting populations reaching differing and suboptimal equilibrium points as they perform a "hill-climbing" style of optimization.
+This analogy has spurred a great deal of research in evolutionary biology, where it has found use in modeling adaptation of alleles, or more recently, in empirical fitness landscapes of proteins.
+The model can in fact be generalized to represent any sort of decision landscape, and has consequently found itself applied broadly in variety of fields such as [Directed evolution, economics?, and the other one].
+Although fitness landscape theory has been used extensively to model adaptation on static genomes or within changing environments, it is less well developed on the scale of long term evolution, where changes in genomic context also arise and can be selected for.
+With this work we extend the Rough Mount Fuji model to incorporate genome evolution alongside allelic mutation to explore the effects of multiple mutational processes on evolutionary outcomes.
 
@@
 
+We examine first the simple case where $\sigma_{e}$ is zero, resulting in a monopeaked landscape with no epistasis.
+Simulations were initialized with the following parameters, and xxx replicates were run by varying the seed value for the mutational rng.
+\begin{center}
+    \begin{tabular}{ | c | c | }
+	\hline
+	Maximum Size & 20 \\ \hline
+	Initial Size & 1, 10, 20 \\ \hline
+	Population & 1000 \\ \hline
+	Generations & 50000 \\ \hline
+	$\sigma_e$ & 0 \\ \hline
+	$\sigma_a$ & 0.1 \\ \hline
+	$\mu$  & 1e-4 \\ \hline
+	$M$ & 1e-4 \\ \hline
+    \end{tabular}
+\end{center}
+
+As there is no epistasis, populations will rapidly reach equilibrium at the fitness maximum on any individual landscape, whereas to find the global peak between all lanscapes a slighly different behavior is observed.
+Since a mutation to $0$ in either the genotype or genome produces an identical fitness, on average $50\%$ of negative additive effects will be removed via each type of mutation, and this balance will gradually fluctuate.
+Figure \ref{additivedynamics} shows genome sizes over time for starting genome sizes 1, 10, and 20 on a landscape with 10 positive additive effects.
+
+\begin{figure}
+	\centering
+	\subfloat[1 locus initially accessible]{\includegraphics[width = 0.4\textwidth]{figures/L20l1i1_20250901_1230_cropped.pdf}}
+	\label{fig:sub1}
+	\subfloat[10 loci initially accessible]{\includegraphics[width = 0.4\textwidth]{figures/L20l10i1_20250901_1229_cropped.pdf}}
+	\label{fig:sub2}
+	\subfloat[20 loci initially accessible]{\includegraphics[width = 0.4\textwidth]{figures/L20l20i1_20250901_1229_cropped.pdf}}
+	\label{fig:sub3}
+	\subfloat[Densities of genome size time series]{\includegraphics[width = 0.4\textwidth]{figures/random_walk_additive.pdf}}
+	\label{additivedynamics:sub4}
+	\caption{caption}
+	\label{additivedynamics}
+\end{figure}
+
+It can be seen that fitness reaches the same value in all three cases within relatively few generations, while genome size simultaneously moves to and thereafter fluctuates about 75\% of the maximum.
+This can be attributed to the random swapping of the two fitness-equivalent mutations as described above, while maintaining a $1$ state for all positive effects.
+In this example, the 10 positive additive effects will always be on, and on average, out of the remaining ten negative additive effects, five will be not present in the genome, while five will be and mutated to a $0$ state, resulting in a mean genome size of 15 averaged over the entire time series (Figure \ref{additivedynamics}d).
+
 \subsection*{Uncorrelated Limit}
@@
 
+As has been noted by studies of the \textit{de novo} evolution of genes, their emergence out of unused genetic substrate remains difficult to observe experimentally.
+(why?
+duplication vs de novo navigation of landscapes).
+
+Owing to a large protein fold space, evolution has generally favored mechanisms of duplication and neofunctionalization amongst new genes through horizontal gene transfer, transposon activity, or whole-genome duplications.
+(by doing so the cell acheives two things: 1.
+relieve pleiotropic effects, 2.
+lessen time to navigate sequence space).
+Alternatively, unused regions of the geome may also see genes spontaneously arise.
+(this process is much slower and harder to observe).
+In either case, these new genes are expected to accumulate neutrally, and over time have the opportunity to gain beneficial functionality.
+Although organisms are effective at removing unnecessary genes through pseudogenization, (this is a slow process, and still does not account for historical contingencies)
+Different organisms may happen to adapt to the same environment in different way
+
+Implications on the expanding genome
+
 \section*{Code and Data Availability}

### 2025-09-01 code/Utils.jl

@@ -383,7 +383,7 @@ function generate_plots(data::DataFrame, μ, M, additive_effects, σ_epi; save::
 
     if save == true
         runtime = Dates.format(now(), "yyyymmdd_HHMM")
-        png("outputs/figures/L$(loci)l$(init_active_loci)i$(max_init_genotype_bits)_$(runtime).png")
+        savefig("outputs/figures/L$(loci)l$(init_active_loci)i$(max_init_genotype_bits)_$(runtime).pdf")
     end
     
     return output_plot

### 2025-09-01 code/additive_analysis.R

@@ -40,15 +40,22 @@ ggsave("output/uniform_additive_t_eq.png", bg = "transparent")
 
 # Now determine if the variable additive landscapes are randomly walking at a 50-25-25% ratio. 
 
-df02 <- read.csv("output/additive_replicate124_data_20250724.csv") %>%
+df02 <- read.csv("outputs/data/04_expanding_additive_replicate6_20250901.csv") %>%
   rename(`1` = X1, `10` = X10, `20` = X20) %>% 
   pivot_longer(cols = c(`1`, `10`, `20`), names_to = "loci")
 
-ggplot(df02, aes(y = value, group = loci, fill = loci)) +
-  geom_histogram(binwidth = 1, position = "dodge") +
+ggplot(df02, aes(y = value, group = loci, fill = loci, color = loci)) +
+  geom_density(bw = 0.5, alpha = 0.5) +
   scale_fill_viridis_d(option = "D", begin = 0.2, end = 0.8) +
-  labs(y = "Genome Size", x = "Count", title = "Histogram of genome sizes (3 reps)") + 
-  geom_hline(yintercept = mean(df02$value), color= "red") +
+  scale_color_viridis_d(option = "D", begin = 0.2, end = 0.8) +
+  ylim(c(0,20)) +
+  labs(y = "Genome Size", 
+       x = "Density", 
+#       title = "Histogram of genome sizes (1e5 Generations)",
+       fill = "Initial genome size",
+       color = "Initial genome size"
+       ) + 
+  geom_hline(yintercept = mean(df02$value), color= "black", lty = 2) +
   theme(plot.title = element_text(hjust = 0.5)) +
   theme_minimal()
 ggsave("output/uniform_additive_rep124.png", bg = "transparent")

### 2025-09-17 docs/thesis_manuscript/manuscript.tex
in the home stretch...

--- 2025-08-28 docs/thesis_manuscript/manuscript.tex
@@
 \usepackage{mathtools}
+%\usepackage{tikz-network}
 \usepackage{graphicx}
@@
 
+%Command for beginning supplementary tables & figures
+\newcommand{\beginsupplement}{
+        \setcounter{table}{0}
+        \renewcommand{\thetable}{S\arabic{table}}%
+        \setcounter{figure}{0}
+        \renewcommand{\thefigure}{S\arabic{figure}}%
+     }
+
 \newcommand{\HRule}[1]{\rule{\linewidth}{#1}}   % Horizontal rule
@@
                         \HRule{2pt} \\ [0.5cm]          % Lower rule + 0.5cm spacing
-                        \normalsize \today                      % Todays date
+%                        \normalsize \today                      % Todays date
                 }
@@
                 \textbf{Jacob Hugh Riina} \\
+		\vspace{2cm}
+		\textbf{2025}
                 \vfill
-                Supervisors\\
-                Prof. Dr. Claudia Bank\\
-                Dr. Stephan Peischl\\
+                Supervisor\\
+                Prof. Dr. C. Bank\\
+                Dr. S. Peischl\\
 }
@@
 
-Since life's origins approximately 3.5 billion years ago, it has generally increased in complexity, and understanding the processes that generate this complexity is one of the main goals of evolutionary biology.
-One such mechanism of generating complexity is through genome evolution: by changing the size of their genomes, the earliest living cells - those with a cell membrane, which encode and pass on information through their DNA - were able to create \textit{evolutionary innovations}: new metabolism, physiology or regulation, which augumented their ability to thrive in a variety of environments.
+Since life's origins approximately 3.5 billion years ago it has generally increased in complexity, and understanding the processes that generate this complexity is one of the main goals of evolutionary biology \cite{duclosInvestigatingEvolutionDevelopment2019}\cite{mcshea2010biology}.
+One such mechanism of generating complexity is through genome evolution \cite{lynchOriginsGenomeComplexity2003}: by changing the size of their genomes, the earliest living cells - those with a cell membrane, which encode and pass on information through their DNA - were able to create \textit{evolutionary innovations}: new metabolism, physiology or regulation, which augumented their ability to thrive in a variety of environments \cite{barveLatentCapacityEvolutionary2013}\cite{uebbingEvolutionaryInnovationsConserved2024}.
 At the increased cost of replicating and proofreading more DNA, organisms have since been able to evolve new genes allowing them to metabolize different carbon, nitrogen, or complex nutrient sources, various types of receptors and secretion systems to interact, competitively or mutualistically, with others, and the internal regulatory networks required for them to decide which behaviors are optimal under which circumstances, or even to coordinate multicellularity.
 Several billion years later, we see the consequences of this incessant evolutionary process: the stunning diversity amongst organisms in the tree of life, each a product of the series of environments they have adapted to.
-We can say, then, that the content of an organism's genome is not solely the result of how it has evolved to fit its current environment, but also the entire history of environments that its ancestors have evolved in; as always, "Nothing in biology makes sense except in the light of evolution".\\
+We can say, then, that the content of an organism's genome is not solely the result of how it has evolved to fit its current environment, but also the entire history of environments that its ancestors have evolved in; as always, "Nothing in biology makes sense except in the light of evolution" \cite{dobzhanskyBIOLOGYMOLECULARORGANISMIC1964}.
 
-Understanding this facet of evolution is [difficult].
-Unlike with [allelic mutation], genome evolution occurs up to several orders of magnitude less freqently, and is more heterogenous in its manifestation.
-We define here genome evolution as a mutational process resulting in either the expansion or reduction in the size of the [functional/coding] genome, stemming from one of several causes: amongst genome-increasing mechanisms, there exist duplications from repeat region replication errors, transposon activity, or whole genome duplications, as well as genes arising from noncoding sequence through de novo gene birth, while genome-reducing mechanisms are pseudogenization and deletions via various replication errors.
-here are some more specifics about the expansion/reduction process.
-here is why they are difficult to observe or create experiments to test.
-here is why is would be beneficial to create models that allow us to ask [] questions that bridge the gap between theory and experiment.
-Owing to a large protein fold space, evolution has generally favored mechanisms of duplication and neofunctionalization amongst new genes through horizontal gene transfer, transposon activity, or whole-genome duplications.
-(by doing so the cell acheives two things: 1.
-relieve pleiotropic effects, 2.
-lessen time to navigate sequence space).
-Alternatively, unused regions of the geome may also see genes spontaneously arise.
-(this process is much slower and harder to observe).
-In either case, these new genes are expected to accumulate neutrally, and over time have the opportunity to gain beneficial functionality.
-Although organisms are effective at removing unnecessary genes through pseudogenization, (this is a slow process, and still does not account for historical contingencies)
-Different organisms may happen to adapt to the same environment in different way
+Understanding this facet of evolution is difficult.
+Unlike with the point mutations that are typically observed in every generation, genome evolution occurs up to several orders of magnitude less frequently \cite{sousaRatesTranspositionEscherichia2013}\cite{lipinskiHighSpontaneousRate2011}, and is more heterogenous in its manifestation.
+We define here genome evolution as a mutational process resulting in either the expansion or reduction in the size of the functional genome, arising from one of several causes: amongst genome-increasing mechanisms, there exist duplications from repeat region replication errors \cite{reamsMechanismsGeneDuplication2015}, transposon activity, or whole genome duplications (WGD), as well as genes arising from noncoding sequence through de novo gene birth \cite{sacconeOriginEvolutionGenes2025}, while genome-reducing mechanisms include pseudogenization and deletions via various types of replication errors \cite{burssedMechanismsStructuralChromosomal2022}.
+Evolution has been observed to utilize both genome reduction \cite{helsenGeneLossPredictably2020}\cite{szameczGenomicLandscapeCompensatory2014} and expansion \cite{vossebergTimingOriginEukaryotic2021} in order to spur adaptation to increase the organism's \textit{fitness}, conceptualized as some metric representing success e.g.
+growth rate within an environment.
+However, more impactful processes such as neofunctionalization of duplications, de novo gene birth, and WGD, remain slow to induce and test, in experiments running for multiple years \cite{tongGenomeDuplicationLongterm2025}.
+There is thus a usefulness to the creation of mathematical models that bridge the gap between theory and experiment and allow us to form testable hypotheses about these phenomena and their impact over long evolutionary timescales.
+One such question naturally arising from the study of this type of adaptive system is regarding the effects of \textit{context dependency} during genome evolution: how could the nonlinear fitness effects of certain genes together nudge evolution in certain directions?
+A good starting point is therefore from mathematical models that allow for the modeling of context, and how it can evolve.
+We thus turn to fitness landscapes.
 
-We do however see in even shorter evolution experiments that not only mutation of alleles, but also genomic abberations appear that explain increases in fitness.
-(noted rate of nucleotide gain per generation stats here)
 \subsection*{Fitness Landscape Theory}
 
-As has been noted by studies of the \textit{de novo} evolution of genes, their emergence out of unused genetic substrate remains difficult to observe experimentally.
-(why?
-duplication vs de novo navigation of landscapes).
-There is thus a usefulness to generation of models that allow us to form hypotheses about these difficult to test systems.
-
-Although fitness landscape theory has been used extensively to model adaptation on static genomes, it is less well developed on the scale of long term evolution, where changes in the genome size also arise.
-With this work we extend the Rough Mount Fuji model to incorporate genome evolution alongside allelic mutation
+The fitness landscape was first introduced in 1932 by Sewall Wright to illustrate the vastness of combinitoric space for alleles in a species \cite{wrightRolesMutationInbreeding}.
+With each unique combination of alleles there can be assigned a fitness value, and so we can generate a topology from a network of possible decision combinations on which we can simulate populations evolving.
+Fitnesses for contexts on the landscape may be higher or lower than expected from the sum of their individual components, leading to "ruggedness", otherwise known as \textit{epistasis}: nonadditivity resulting from higher-order interactions between components, that ultimately leads to, in the case of evolution, adapting populations reaching differing and suboptimal equilibrium points as they perform a "hill-climbing" optimization.
+This analogy has spurred a great deal of research in evolutionary biology, where it has found use in modeling macro-level adaptation \cite{venkataramMutualismenhancingMutationsDominate2023}\cite{frankeEvolutionaryAccessibilityMutational2011}, or more recently, in empirical fitness landscapes of proteins \cite{papkouRuggedEasilyNavigable2023}.
+The model can in fact be generalized to represent any sort of decision landscape, and has consequently found itself applied broadly in variety of fields such as directed evolution \cite{alpayEffectsSelectionStringency2024}, theoretical economics \cite{khraishaComplexEconomicProblems2020}, or management studies \cite{siggelkowFirmsSystemsInterdependent2011}.
+Although fitness landscape theory has been used extensively to model adaptation on static genomes or within changing environments, it is less well developed on the scale of long term evolution, where changes in genomic context also arise and can be selected for.
+With this work we extend the Rough Mount Fuji model to incorporate an additional mechanism of genome evolution alongside allelic mutation to explore the effects of multiple mutational processes on evolutionary outcomes.
 
@@
 
-\subsection*{The Rough Mount Fuji Model}
+\subsection*{A Classical Rough Mount Fuji Model}
 
-The Rough Mount Fuji (RMF) model allows for tunable epistasis \cite{neidhartAdaptationTunablyRugged2014} and so enables the investigation its effects as landscapes scale.
-In this model we represent genotypes as binary vectors, and their permutations in genotype space construct a network where each has a hamming distance of 1 from its neighbors, and the degree of each node is equivalent to the number of loci $L$ in the model (see Figure~\ref{fig:genotype_space}).
+The Rough Mount Fuji (RMF) model allows for tunable epistasis \cite{neidhartAdaptationTunablyRugged2014}.
+In the standard model, we represent genotypes as binary vectors, and their permutations in genotype space construct a network where each has a hamming distance of 1 from its neighbors, with the degree of each node equivalent to the number of loci $L$ in the model (see Figure \ref{genotype_space}).
 
-The assignment of fitness values to genotypes entails sampling two probabilistically determined components: Locus-wise additive effects $a = (a_1, \cdots, a_i) \sim \text{iid.
-} \mathcal{N}(0, 1)$ are drawn from a normal distribution, and each genotype $g$ recieves a unique epistatic component $e_g \sim \mathcal{N}(0, \sigma_e)$, also drawn from a normal distribution.
-Fitness values for some arbitrary genotype can then be assigned as the sum of the product of the additive effects and the genotype, plus the epistatic component, as an exponential: $$f_g = exp{(e_g + \sum_{i=0}^{L} a_i * g_i)}$$ Epistasis is represented in the ratio between the variance of additive components and epistatic components, and we control epistasis on the landscape by varying the epistatic $\sigma_e$ - in cases where the epistatic component is small the additive component dominates and the landscape becomes easily predictable, whereas when epistatic effects are large and outweigh the additive component, there is virtually no correlation between different genotypes and ruggedness is high.
-Now able to reproducibly retrieve some epistatic component for any genotype by providing the CBRNG with the base-10 representation of that binary genotype, we are no longer obligated to generate fitnesses for all genotypes on the landscape in advance and instead need only do so upon encountering that genotype during simulation.
+\begin{figure}[h!]
+	\centering
+	\includegraphics[width = 0.5\textwidth]{figures/simple_landscape.png}
+	\caption{A 4-dimensional hypercube representing a 4-locus model.
+Genotypes are represented as binary strings e.g.
+[0000], and evolution generally moves over the landscape from low (blue) to high (green) fitness.}
+	\label{genotype_space}
+\end{figure}
 
-A discrete-generation RMF model with Wright-Fisher dynamics was implemented based on \cite{liRapidAdaptationRecombining2024} and previous software \cite{amadoSTUNForwardtimeSimulation2023}.
-We model a static $N$ total individuals mutating and reproducing simultaneously in each generation, with genotype frequencies governed by binomial and multinomial sampling, respectively.
-To initialize the simulation, we specify the number of loci in the model and consequently the length of the vector of additive effects to generate, the number of randomly distributed initial genotypes on the landscape and the number of individuals per genotype, the simulation length, the variance $\sigma$ of the epistatic component, and the mutation rate $\mu$ - the probability for an individual to generate a mutant of hamming distance 1 from its parent.
-An initial report table is then generated with a description of initial genotypes on the landscape, their population sizes, and fitnesses.
-We proceed to instantiate subsequent generations by first generating mutants from each extant genotype with population $\mathbb{P}$ through binomial sampling: $\text{Binomial}(n = \mathbb{P}, p = \mu)$.
-We adjust the corresponding population counts and calculate fitnesses for new genotypes if necessary, and then assign from $i$ extant genotypes for all individuals in the new generation of size $N$ new genotypes according to relative fitness: $\text{Multinomial}(N, p^{\prime} = (p^{\prime}_1, \cdots, p^{\prime}_i))$ given $p^{\prime}_i = p_i * \frac{\omega_i}{\bar{\omega}}$, where $p_i$ is some genotype's current frequency, $\omega_i$ is its fitness, and $\bar{\omega}$ is the average fitness across all extant genotypes.
+The assignment of fitness values to genotypes entails sampling two probabilistically determined components: Locus-wise additive effects $A = (a_1, \cdots, a_i) \overset{iid}{\sim} \mathcal{N}(0, \sigma_a)$ are drawn from a normal distribution, and each genotype $g$ recieves a unique epistatic component $e_g \sim \mathcal{N}(0, \sigma_e)$.
+Fitness values for some arbitrary genotype can then be assigned as the sum of the product of the additive effects and the genotype, plus the epistatic component, as an exponential:
+
+$$f_g = \exp{\left(e_g + \sum_{i=0}^{L} a_i g_i\right)}$$
+
+In this way, a 0 in the genotype represents a "neutral" allele, and 1 an allele with either a positive or negative effect.
+Epistasis is represented in the ratio between the variance of additive components and epistatic components, and we control its effect on the landscape by varying the epistatic $\sigma_e$ relative to $\sigma_a$ - in cases where the epistatic component is small the additive component dominates and the landscape becomes easily predictable, whereas when epistatic effects are large and outweigh the additive component, there is virtually no correlation between different genotypes and ruggedness is high.
+We proceed to model populations of individuals with different genotypes adapting on these landscapes through numerical simulations.
+
+A discrete-generation RMF model with Wright-Fisher dynamics was implemented based on Li et al.
+2024 \cite{liRapidAdaptationRecombining2024} and previous software \cite{amadoSTUNForwardtimeSimulation2023}.
+We model a static $N$ total individuals mutating and reproducing simultaneously in each generation, where each individual may be part of one of many subpopulations of genotypes $g$.
+To initialize the simulation, we specify the maximum number of loci in the model and consequently the length of the vector of additive effects to generate, the simulation length, the variance $\sigma_e$ of the epistatic component, and the mutation rate $\mu$ - the probability for an individual to generate a mutant of hamming distance 1 from its parent.
+We proceed to instantiate subsequent generations by first generating mutants from each extant genotype with population $P$ through binomial sampling: $\text{Binomial}(n = P, p = \mu)$.
+We adjust the corresponding subpopulation counts for each genotype and calculate fitnesses for new genotypes if necessary, and then assign from $i$ extant genotypes for all individuals in the new generation of size $N$ new genotypes according to their relative fitness: $\text{Multinomial}(N, p^{\prime} = (p^{\prime}_1, \cdots, p^{\prime}_i))$ given $p^{\prime}_i = p_i * \frac{\omega_i}{\bar{\omega}}$, where $p_i$ is some genotype's current frequency, $\omega_i$ is its fitness, and $\bar{\omega}$ is the average fitness across all extant genotypes.
 Therefore, highly fit mutants gain an advantage during sampling and are less likely to be eliminated through drift.
-The simulation is run for the specified number of generations and upon terminating a final report is also provided (Figure~\ref{fig:report}).
+Simulations were generally run for 50000 generations before terminating.
 
@@
 
-The facile generation of random epistatic components was performed using a Julia implementation \cite{JuliaRandomRandom123jl2024} of the Random123 package from D.E.
-Shaw Research \cite{salmonParallelRandomNumbers2011}.
-Particularly, the ARS1x generator was used, with a period of $2^{128}$, this being the number of loci that can be modeled in a genotype network without reseeding.
+With this manner of simulation, the same genotype may arise and disappear multiple times due to drift - we then require the ability to reproducibly generate many unique epistatic components to calculate the fitness of any possible genotype on the landscape.
+We typically do this by mapping a sequence of random numbers in the state space of a pseudorandom number generator (PRNG) to base-10 genotypes, but as the size of genotype space grows large, retrieval in $O(n)$ time becomes unwieldy.
+Figure \ref{prng} shows the process of generating random numbers with a PRNG, where in order to generate the $i^{\text{th}}$ random number $r$ in a sequence, $i-1$ prior values must be calculated.
+To mitigate this problem, we utilize counter-based random number generators (CBRNGs), which can retrieve any number from the state space of the generator in constant time as a function of a seed number and an index (Figure \ref{cbrng}).
+In our implementation, the generation of random epistatic components was performed in Julia using an ARS1x generator \cite{JuliaRandomRandom123jl2024} of the Random123 package from D.E.
+Shaw Research \cite{salmonParallelRandomNumbers2011}, with a state space of $2^{128}$, this being the number of genotypes that can be modeled in a network.
+
+\begin{figure}[h!]
+	\centering
+	\subfloat[\label{prng} PRNG]{\includegraphics[width = 0.75\textwidth]{figures/prng.pdf}}
+	\\
+	\subfloat[\label{cbrng} CBRNG]{\includegraphics[width = 0.75\textwidth]{figures/cbrng.pdf}}
+	\label{rngs}
+	\caption{\textit{(a)} Scheme for a PRNG: each random number $r_i$ in the generator yields the next state $r_{i+1}$ through $f(r_i)$.
+Each random number $r_i$ is mapped to a base 2 number in genotype space based on its index $i$.
+Generators have a state space of $\mathcal{S}$ values, after which the values output will cycle back to the seed.
+\textit{(b)} Scheme for a CBRNG: Each state $r_i$ in the generator can be yielded in constant time from only the seed sequence and an index.
+These will also cycle, when inputting an index larger than $\mathcal{S}$.}
+\end{figure}
+
+\subsection*{A Model Extension for Genome Evolution}
+
+To allow for genome size changes, we extend the model to give individuals a genome $G$ and a genotype $g$ on that genome, and calculate fitness as a function of both.
+Each $G$ and $g$ are binary strings of the maximum possible length for the genome ($L_{\text{max}}$), and we utilise $G$ to determine which loci in $g$ may mutate: a 1 value in the genome represents the presence of a gene or functional genetic element, and allows the corresponding position in $g$ to be mutated between 0 and 1, while a 0 in $G$ ensures that position in $g$ will always be 0, and thus never contribute additively to fitness.
+Both mutational processes (genome and genotype mutation) are assigned their own mutation rate parameter ($M$ and $\mu$, respectively), and therefore to access the highest fitness states, populations must first select for genomic changes in $G$ that allow them to mutate their genotypes $g$ to those fitness peaks.
+Additionally, we assume that when new genes evolve, they may interact nonadditively with other previously existing genes, and thus genomic changes in $G$ must also be able to change the epistatic component, not only genotype $g$ changes: we therefore make the epistatic component $e_{G,g}$.
+As the CBRNG function takes both a seed and an index, we use the genome as a seed and the genotype as the index for to retrieve unique epistatic components for every combination of genome and genotype in 128-bit space.
+Effectively, we allow for individuals to navigate a network of landscapes of different sizes, simultaneously adapting on those landscapes and transitioning between them by changing their genomes.
 
 \subsection*{Scripting}
+
+All functions for the simulation of this model were written in a `Utils.jl` file.
+This includes methods for the generation of additive effects, the generation of mutants from a population, or the resampling of a population for a new generation.
+To allow for reproducibility, the model requires seeds for PRNGs that determine the initial genome and genotype, the additive effects vector, and the mutations made in each generation.
+In this way, rerunning a simulation with identical parameters but varying only the mutation seed allows us to simulate replicates on the same landscape and starting position, that may generate different mutants and thus different outcomes over the course of the simulation.
+Examinations of different classes of landscapes was then performed in separate scripts, each by defining different parameters for the model, and for each replicate, initializing the PRNGs with the appropriate seeds, and running the simulation function.
 
@@
 
+We examine first the simple case where $\sigma_{e}$ is zero and the values of the additive effects $a_i \neq 0$, resulting in a monopeaked landscape with no epistasis.
+Simulations were initialized with the following parameters, and 100 replicates were run by varying the seed value for the mutational rng.
+
+\begin{center}
+    \begin{tabular}{ | c | c | }
+	\hline
+	Maximum Size & 20 \\ \hline
+	Initial Size & 1, 10, 20 \\ \hline
+	Population & 1000 \\ \hline
+	Generations & 50000 \\ \hline
+	$\sigma_e$ & 0 \\ \hline
+	A & $a_i \overset{iid}{\sim} N(\mu_a = 0, \sigma_a = 0.1)$ \\ \hline
+	$\mu$  & 1e-4 \\ \hline
+	$M$ & 1e-4 \\ \hline
+    \end{tabular}
+\end{center}
+
+As there is no epistasis, populations will rapidly reach equilibrium at the fitness maximum on any individual landscape, whereas to find the global peak between all lanscapes a slighly different behavior is observed.
+Since a mutation to $0$ in either the genotype or genome produces an identical fitness, on average $50\%$ of negative additive effects will be removed via each type of mutation, resulting in half of them being present on the genome with $G_i = 1, g_i = 0$, and half not present with $G_i = g_i = 0$.
+This balance will gradually fluctuate.
+Figure \ref{additivedynamics} shows genome sizes over time for starting genome sizes 1, 10, and 20 for one replicate with 10 positive additive effects.
+
+\begin{figure}[h!]
+	\centering
+	\subfloat[\label{ad:sub1} Initial genome size 1]{\includegraphics[width = 0.33\textwidth]{figures/L20l1i1_20250902_1035.pdf}}
+	\subfloat[\label{ad:sub2} Initial genome size 10]{\includegraphics[width = 0.33\textwidth]{figures/L20l10i1_20250902_1035.pdf}}
+	\\
+	\subfloat[\label{ad:sub3} Initial genome size 20]{\includegraphics[width = 0.33\textwidth]{figures/L20l20i1_20250902_1035.pdf}}
+	\subfloat[\label{ad:sub4} Densities of genome size time series]{\includegraphics[width = 0.33\textwidth]{figures/random_walk_additive.pdf}}
+	\caption{\textit{(a), (b), (c)}: Time series of genome size and fitness for example simulations starting at 1, 10, and 20 loci.
+\textit{(d)}: densities of genome size time series all form a mean around 15.}
+	\label{additivedynamics}
+\end{figure}
+
+It can be observed that fitness reaches the same value and plateaus in all three cases within relatively few generations, while genome size simultaneously moves to and thereafter fluctuates about 75\% of the maximum.
+This can be attributed to the random swapping of the two fitness-equivalent mutations as described above, while maintaining a $1$ state for all positive effects.
+In this instance, the 10 positive additive effects will always be $G_i = g_i = 1$, and on average, out of the remaining ten negative additive effects, five will be not present in the genome ($G_i = g_i = 0$), while five will be and mutated to a $0$ state ($G_i = 1, g_i = 0$), resulting in a mean genome size of 15 averaged over the entire time series (Figure \ref{ad:sub4}).
+We can see how, in the complete absence of epistasis, there are many landscapes that produce equivalent optimal fitness and as such there is no single equilibrium point, and that populations are not constrained from reaching these fitness-optimal states by their starting genome size.
+
 \subsection*{Uncorrelated Limit}
 
-\subsection*{Time-Dynamic Outputs}
+We next investigate uncorrelated (known alternatively as House of Cards) landscapes, where all additive effects $a_i = 0$ and $\sigma_e$ is positive.
+This produces maximally rugged landscapes under the RMF model, where each genotype on a landscape and each landscape in genome space has a randomly drawn fitness and is uncorrelated to any other.
+We simulated 200 replicates under the following conditions:
 
-\subsection*{Endpoint Outputs}
+\begin{center}
+    \begin{tabular}{ | c | c | c | }
+        \hline
+	Maximum Size & 20 & 100 \\ \hline
+	Initial Size & 1, 5, 10, 15, 20 & 1, 25, 50, 75, 100 \\ \hline
+	Population & 1000 & " \\ \hline
+	Generations & 50000 & " \\ \hline
+	$\sigma_e$ & 1 & " \\ \hline
+	$A$ & All $a_i = 0$ & " \\ \hline
+	$\mu$  & 1e-4 & " \\ \hline
+	$M$ & 1e-4 & " \\ \hline
+    \end{tabular}
+\end{center}
 
-Does endpoint exist under this model?
+As fitnesses are determined solely through the sampling of the epistatic component distribution $\epsilon_{G,g} \overset{iid}{\sim} N(0,\sigma_e)$, larger landscapes will tend to have higher fitness peaks overall, due to a greater number of samples producing, on average, more extreme values.
+This effect can be shown analytically in the expectation of the distribution of fitness peaks, as determined through extreme value theory:  $$\mathbb{E}[X] = \exp{\left(\sqrt{2L \ast (\frac{L \ast \sigma_a^2}{2} + \sigma_e^2) \ast \ln(2)}\right)}$$ Given a constant variance of additive and epistatic terms between landscapes, the mean value of fitness peaks increases with the number of loci as an exponent of $e$, either sub- or superlinearly depending on parameters (Supplementary figure \ref{supp:peak_scaling}).
+We see this confirmed in Figures \ref{ul:sub1} - \ref{ul:sub3}, where in both static (genome mutation parameter $M = 0$) and expanding ($M < 0$) landscape cases, an increased starting number of loci leads on average to higher fitness values.
+We also observe when comparing initial genome size 1 between expanding landscapes with $L_{\text{max}}$ of 20 and 100, that final fitnesses also scale with maximum size.
+However, final genome sizes do not appear to reach these sizes associated with higher fitness maxima; they are instead highly correlated with initial size.
+Figures \ref{ul:sub4} and \ref{ul:sub5} visualize the distributions of final genome sizes by initial size, wherein genome sizes end in a relatively compact distribution not far from the initial size.
+These distributions appear to skew in the direction of half the maximum genome size, and the magnitude of such remains constant as the number of loci increases.
+At the extremal case of $L_{\text{initial}} = 1$, for both $L_{\text{max}} = 20$ and $L_{\text{max}} = 100$ the skew of the final size distribution is around 5, and correspondingly for test cases at $^1/_4$, $^1/_2$, and $^3/_4$ of the maximum genome size the skew is proportionally similar regardless of $L_{\text{max}}$.
+This effect appears to be a mutational bias from the genome size limit, where smaller genomes that mutate are by chance more likely to change a 0 in their genome to a 1, and conversely a large genome with many 1s is more likely to reduce its genome size by changing to a 0.
+We conclude that larger genomes will have higher fitness outcomes on average, yet see that on landscapes with epistasis populations do not indefinitely increase their genome size - there is a finite skew to the final sizes that constrains achievable fitness.
+
+\begin{figure}[h!]
+	\centering
+	\subfloat[\label{ul:sub1} Final fitnesses:\\ Static landscapes]{\includegraphics[width = 0.333\textwidth]{figures/05-2_static_finalfitness.pdf}}
+	\subfloat[\label{ul:sub2} Final fitnesses:\\ Expanding landscapes \\($L_{\text{max}} = 20$)]{\includegraphics[width = 0.333\textwidth]{figures/06-1_expanding_initialsize_fitness.pdf}}
+	\subfloat[\label{ul:sub3} Final fitnesses:\\ Expanding landscapes \\($L_{\text{max}} = 100$)]{\includegraphics[width = 0.333\textwidth]{figures/06-4_expanding_initialsizefitness_2.pdf}} \\
+	\subfloat[\label{ul:sub4} Final size distributions \\($L_{\text{max}} = 20$)]{\includegraphics[width = 0.333\textwidth]{figures/06-3_expanding_initialfinalsize.pdf}}
+	\subfloat[\label{ul:sub5} Final size distributions \\($L_{\text{max}} = 100$)]{\includegraphics[width = 0.333\textwidth]{figures/06-6_initialfinalsize_2.pdf}}
+	\caption{Final fitnesses of \textit{(a)} static and \textit{(b),(c)} expanding landscapes, \textit{(d),(e)} final sizes of expanding landscapes.}
+	\label{uncorrelated_fitnesses}
+\end{figure}
+
+\pagebreak
+
+\subsection*{Intermediate Epistasis}
+
+We last turn to rugged landscapes with both additive and epistatic components, which are most similar to empirical landscapes. We ran 5000 replicates of simulations with the following parameters:
+
+\begin{center}
+    \begin{tabular}{ | c | c | }
+	\hline
+	Maximum Size & 20 \\ \hline
+	Initial Size & 5 \\ \hline
+	Population & 1000 \\ \hline
+	Generations & 50000 \\ \hline
+	$\sigma_e$ & 0.1 \\ \hline
+	$A$ &  $a_i \overset{iid}{\sim} N(\mu_a = 0, \sigma_a = 0.1)$ \\ \hline
+	$\mu$  & 1e-4 \\ \hline
+	$M$ & 1e-5 \\ \hline
+    \end{tabular}
+\end{center}
+
+By setting the genome mutation rate $M$ an order of magnitude less than the genotype mutation rate $\mu$, we separate the time scales on which peaks on a landscape and transitions between landscapes are navigated to, allowing fitness peaks to be reached on each landscape before genome mutations are made.
+We track the genomic trajectories of these simulations over time, observing which final genome sizes are achieved, and the fitness peaks reached on each intermediate landscape along the way.
+
+Figure \ref{ie:sub1} visualizes the genome size trajectories of all simulations as they increase from an initial size of five until the end of the simulation.
+To simplify the plots, we remove replicates that do not have a monotonically increasing genome size, although the effect in these cases is consistent with the rest of the data (Supplementary figure \ref{uf}).
+We observe that although populations must not necessarily take fitness increasing paths to reach a peak on any given landscape, when transitioning between landscapes there does seem to be consistent fitness increases.
+Additionally, replicates that are able to increase their genome size seem to occupy lower fitness regions of each landscape, as opposed to those that end the simulation at some landscape size, which tend to have higher fitness - this behavior is further examined in figure \ref{ie:sub2}.
+Here we see the distributions of fitness values at each genome size split between those replicates which ended the simulation at that size, versus those that continued to expand their genome.
+Consistently among all genome sizes, we notice that replicates that increase their genome size past some size actually have, on average, lower fitnesses that those that do not.
+This implies that populations that begin less fit on some landscape and who minimize their fitness gain on each subsequent landscape possess a greater ability to expand their genome.
+
+\begin{figure}[h!]
+        \centering                                                                                                                                                  \subfloat[\label{ie:sub1} Fitness trajectories]{\includegraphics[width = 0.4\textwidth]{figures/genome_trajectories.pdf}}
+        \subfloat[\label{ie:sub2} Endpoints]{\includegraphics[width = 0.4\textwidth]{figures/endpoint_boxplots.pdf}} \\                                             \subfloat[\label{ie:sub3} Size transition timings]{\includegraphics[width = 0.4\textwidth]{figures/timings.pdf}}                                            \subfloat[\label{ie:sub4} Endpoint comparisons on uncorrelated landscapes.]{\includegraphics[width = 0.4\textwidth]{figures/hoc_boxplots.pdf}}              \caption{\textit{(a)}: Fitness trajectories over different genome sizes, \textit{(b)}: fitness comparisons of endpoint and non-endpoint simulations at various genome sizes, \textit{(c)}: Distributions of generation times at which genome size transitions are made, \textit{(d)}: fitness comparisons as in \textit{(b)} on uncorrelated landscapes}
+        \label{intermediate_epistasis}
+\end{figure}
+
+\pagebreak
 
@@
 
+	[The results of out simulations reveal some of the new dynamics we observe on expanding landscapes.
+We see how epistasis not only constrains navidability on individual landscapes but the transition between them as well.
+With additive landscapes, there is nothing preventing the collection of all possible additive effects regardless of starting point, whereas when epistasis increases, we see an increased amount of correlation between the starting and ending genome size.
+Most interestingly, we see what happens regarding adaptation over multiple landscapes, who gets the highest fitnesses?
+It's actually the ones that start from lower fitness and minimize fitness gain between when increaseing size that end up with higher fitness.
+We think this is due to an interaction between the scaling of the fitness landscape peak distribution and [xyz].
+As you get fitter, this decreases the probability that any of your neighborw will yield a fitness increasing path, and this applies when jumping between landscapes as well.
+There fore adapting well (to a higher fitness peak) decreases your ability to increase genome size (there is some formula for the uncorrelated landscacpe that shows this).
+However, we also know from examining final fitnesses achieved on larger landscapes, that the mean of the distribution increases, because there are more states in genotype space and therefore more extreme fitnesses have the potential to be generated (This is also sensible.
+The larger the solution space, ther more liekely vert goodsolutions will be found).
+We can then state these two things succinctly:
+	\begin{enumerate}
+		\item Larger genome size results on average in higher fitnesses.
+
+		\item Higher fitness reduces ability to increase genome size.
+	\end{enumerate}
+	These contradictory objectives in the context of epistasis result in the behavior that we see.
+Assuming a low enough genome mutation rate, fitenss peaks will be reached in the current landscape before landscapes will change.
+Thus we can claim that those that hit lower fitness peaks on any landscape, have a greater ability to increase their genome size, and therefore have a higher end probability to have among higher fitnesses, while never having a particularly high fitness (relative to other paths) over the course of the system's evolution.
+]
+
+\subsection*{A model for the evolution of metabolism}
+
+This is very intriguing.
+It suggests that systems that optimize and encode heritable information can actually benefit from avoiding high fitness states too early in the evolution of the system, and finds itself in concordance with results from a similar long-term evolution experiment.
+\cite{woodsSecondOrderSelectionEvolvability2011} To illustrate a biologically relevant scenario in which this would apply, we here introduce a toy model for the evolution of a metabolic network.
+
+\begin{figure}[h!]
+	\centering
+	\includegraphics[width = 0.33\textwidth]{figures/evolving_metabolism.pdf}
+	\caption{Example biochemical network space for a genome of maximum size 7.}
+	\label{evomet}
+\end{figure}
+
+Here we represent chemical species as nodes and enzymatic reactions ($r_1\cdots r_i$) as edges with catalytic rate weights ($w_1 \cdots w_i$), and a genome can be comprised of any combination of genes for these enzymes.
+As shown in Figure \ref{evomet}, a typical cell will take nutrients (red, source nodes), and metabolize them through a series of biochemical intermediates (yellow nodes) to growth metabolites (orange, sink nodes) that cell requires to grow and divide, such as amino acids, nucleotides, or phospholipids.
+We therefore consider a metabolically viable cell to be one that has genes in its genome that construct at least one unbroken path between a source node and sink node.
+Fitness (or flux) calculations for larger, more empirical network models are typically done via flux balance analysis, \cite{orthWhatFluxBalance2010} but we will instead assume that the fitness for some genome is similar to that of the RMF model: $$f_G = e_G + \displaystyle\sum_{i=0}^{L}r_i w_i $$ That is, fitness is the sum of all weights $w_i$ for reactions of enzymes present in the genome plus an epistatic term $e_g$ to allow for tunable nonadditivity of a particular reaction set, with one additional rule: only reactions which lie directly on a source-sink path contribute to fitness.
+Like in the RMF model, we can then sample $e_G$ from $X \overset{iid}{\sim} \mathcal{N}(\mu, \sigma)$ and weights $w_i$ from a distribution with support between 0 and infinity such as a gamma, exponential, or absolute-value normal, as we assume enzyme rates in this context are generally not negative, nor catalytically bidirectional.
+
+This creates for us a rugged landscape with fitness maxima, or metabolically viable subnetworks, determined by the topology of the biochemical space chosen.
+We will not examine the consequences of random topologies on the outcomes of this model, but focus on how simply one trivially determined topology can produce behavior consistent with the previous results on expanding landscapes.
+At initiation, a genome size of three is the first point at which there are multiple viable genomes, encoding two different pathways for the metabolism of some nutrient - in this example, they can be represented as $1110000$ (or a reaction set of $[r_1, r_2, r_3]$) and $0001110$ ($[r_4, r_5, r_6]$).
+This is already quite common in bacterial physiology, where for example certain species may use the Embden-Meyerhof-Parnas \cite{sanchez-pascualaRefactoringEmbdenMeyerhof2017} or the Entner-Doudoroff \cite{conwayEntnerDoudoroffPathwayHistory1992} glycolysis pathways to produce the same products.
+All other factors equal, each pathway will have higher fitness in half of all cases, with the magnitude of the difference between the two a function of the variance of additive and epistatic components, and the current genome size.
+Then, upon expansion of the genome, one pathway immediately gains access to another nutrient, while the other must take a longer path in order to access the same - this may be something present in the environment that had been underutilized, and represents niche expansion.
+On a purely additive landscape, this is no problem for either starting point.
+Since the optimal solution is to possess all genes in metabolic space and adding them is neutral, so we would see both starting points expand their genomes to the maximum size, similarly to how the additive expanding landscapes were all able to reach the optimal equilibrium size.
+However, once introducing epistasis, the ability to add genes is constrained by the probability the the epistatic component increases with every consecutive genome expansion, and the greater it is at the beginning, or the larger the number of genes to add to complete a pathway, the less likely this is to be the case.
+Provided that the variance of reaction weights is sufficiently low relative to their mean, this will commonly cause the lower fitness variant at size three to gain higher fitness at size four, while the alternative pathway will likely not be able to expand its genome to also utilize the new resource.
+This also echoes the behavior seen on the expanding landscape models - through the evolution of the system to larger sizes, epistasis begins to constrain free genome expansion, making it instead highly contingent on initial size, and additionally gives an advantage to lower fitness variants to expand their genomes.
+
+[What is the significance of this work?
+Here we show a new model that incorporates an infrequently modeled part of evolution, that the genome can fluctuate both in content and size, and that the behavior seen in this expanded model is not a simple recapitulation of a simpler fitness landscape.
+This matters because I want to demonstrate that due to the darwinian nature of evolution, those lower fitness paths are less likely to be taken in favor of higher fitnesses - this is the consequence of competition.
+Therefore, I would hypothesize that a non-negligible amount of the particular forms life takes now can be attributed to this process.
+I hope that these findings and the model itself allow researchers to  ]
+
 \section*{Code and Data Availability}
+
+All code, figures, and associated datasets may be accessed at \url{www.github.com/riinajh/expanding_landscapes}
+
+\section*{Acknowledgements}
+
+Special thanks to Suman G. Das for his analytical solution of the fitness peak distribution mean amd the rest of the THEE lab for their valueable comments and critiques throughout this work.
 
@@
 
+\section*{Supplementary Material}
+
+\beginsupplement
+
+\begin{figure}[h!]
+	\centering
+	\includegraphics[width = 0.5\textwidth]{figures/mean_peak_scaling.pdf}
+	\caption{}
+	\label{supp:peak_scaling}
+\end{figure}
+
+\begin{figure}[h!]
+	\centering
+	\subfloat[\label{uf:gt} Fitness trajectories for all simulations, not filtered for genome increasing replicates]{\includegraphics[width = 0.4\textwidth]{figures/genome_trajectories_unfiltered.pdf}}
+	\subfloat[\label{uf:bp} Fitness comparisions of endpoint and non-endpoint simulations, unfiltered]{\includegraphics[width = 0.4\textwidth]{figures/endpoint_boxplots_unfiltered.pdf}}
+	\caption{}
+	\label{uf}
+\end{figure}
+
 \end{document}

### 2025-09-17 code/additive_analysis.R

@@ -36,7 +36,7 @@ ggplot(data, aes(x = loci_diff, y = as.factor(timepoint), fill = as.factor(Loci)
        color = "Loci") +
   theme(plot.title = element_text(hjust = 0.5)) +
   theme_minimal()
-ggsave("output/uniform_additive_t_eq.png", bg = "transparent")
+ggsave("outputs/figures/uniform_additive_t_eq.pdf", bg = "transparent")
 
 # Now determine if the variable additive landscapes are randomly walking at a 50-25-25% ratio. 
 
@@ -58,4 +58,4 @@ ggplot(df02, aes(y = value, group = loci, fill = loci, color = loci)) +
   geom_hline(yintercept = mean(df02$value), color= "black", lty = 2) +
   theme(plot.title = element_text(hjust = 0.5)) +
   theme_minimal()
-ggsave("output/uniform_additive_rep124.png", bg = "transparent")
+ggsave("outputs/figures/additive_densities_rep6.pdf", bg = "transparent")

### 2025-09-17 code/epistatic_analysis.R

@@ -7,6 +7,15 @@ setwd("~/mastersarbeit/thesis/code")
 expected_max_fitness <- function(L, add, epi){
   return(exp((2 * L * (((L / 2) * add^2) + epi^2) * log(2))^0.5))
 }
+Z <- expand_grid(Loci = 1:20, sig_a = c(0, 0.1), sig_e = c(0, 0.1)) %>% 
+  mutate(mean_max = expected_max_fitness(Loci, sig_a, sig_e)) %>% 
+  filter(!(sig_a == 0 & sig_e == 0))
+ggplot(Z, aes(x = Loci, y = mean_max)) +
+  facet_grid(rows = vars(sig_a), cols = vars(sig_e)) +
+  geom_point() +
+  labs(y = "Mean Peak Fitness") +
+  theme_minimal()
+ggsave("outputs/figures/mean_peak_scaling.pdf")
 
 ################################################################################
 # Begin with the static landscape
@@ -31,18 +40,18 @@ ggsave("outputs/figures/05-1_static_timeseries.png")
 
 # Also plotting the final fitnesses vs size for comparison
 # Filter only for endpoints as to compare to later simulations
-ggplot(filter(df01, Generation == 50000), aes(x = as.factor(Loci), y = Final_fitness, group = Loci,
+ggplot(filter(df01, Generation == 50000), aes(x = as.factor(Loci), y = log10(Final_fitness), group = Loci,
                                               fill = as.factor(Loci), color = as.factor(Loci))) +
   geom_boxplot() +
-  lims(y = c(0,200)) +
+  lims(y = c(0,2.5)) +
   scale_fill_viridis_d(option = "B", begin = 0.25, end = 0.9, alpha = 0.7) +
   scale_color_viridis_d(option = "B", begin = 0.25, end = 0.9) + 
-  labs(x = "Genome Size", y = "Fitness",
-       title = "Final Fitnesses of Static Uncorrelated Landscapes (99 Replicates)",
-       fill = "Genome Size (Fixed)",
-       color = "Genome Size (Fixed)") +
+  labs(x = "Genome Size", y = "Log Fitness",
+       #title = "Final Fitnesses of Static Uncorrelated Landscapes (99 Replicates)",
+       fill = "Genome Size (Static)",
+       color = "Genome Size (Static)") +
   theme_minimal()
-ggsave("outputs/figures/05-2_static_finalfitness.png")
+ggsave("outputs/figures/05-2_static_finalfitness.pdf")
 
[…]

### 2025-09-17 code/intermediate_analysis.R

@@ -1,6 +1,5 @@
 library(tidyverse)
 library(ggridges)
-library(ggpmisc)
 setwd("~/mastersarbeit/thesis/code")
 
 # Function for the solution from Suman that yields the expectation for the distribution
@@ -8,14 +7,6 @@ setwd("~/mastersarbeit/thesis/code")
 expected_max_fitness <- function(L, add, epi){
   return(exp((2 * L * (((L / 2) * add^2) + epi^2) * log(2))^0.5))
 }
-
-################################################################################
-# Begin with the static landscape
-################################################################################
-
-filepath_df01 <- "outputs/data/"
-df01 <- read.csv(filepath_df01) %>% 
-  mutate("percent_max" = Final_fitness * 100 / expected_max_fitness(Loci, 0, 1))
 
 ################################################################################
 # Now the expanding landscape
@@ -34,66 +25,149 @@ ggplot(df02, aes(x = Final_genome_size)) +
 
 # 7, 8, 9 are all good candidates. 10 will let us analyze the nonchanging genomes as well.
 
-ggplot(filter(df02, Final_genome_size == 6), aes(x = as.factor(Loci), y = Final_fitness, group = as.factor(Loci))) +
+ggplot(filter(df02, Final_genome_size == 7), aes(x = as.factor(Loci), y = Final_fitness, group = as.factor(Loci))) +
   geom_boxplot()
 
 ################################################################################
 # Looking at dynamics (through the series of selective sweeps) of expanding landscapes
 ################################################################################
 
-filepath_df03 <- "outputs/data/10_dynamics_intermediate_expanding_landscape_20250828_1702.csv"
-df03 <- read.csv(filepath_df03) %>% 
+filepath_df03.1 <- "outputs/data/10_dynamics_intermediate_expanding_landscape_20250828_1702_1kreps.csv"
+filepath_df03.2 <- "outputs/data/10_dynamics_intermediate_expanding_landscape_20250831_2329_2kreps.csv"
+filepath_df03.3 <- "outputs/data/10_dynamics_intermediate_expanding_landscape_20250903_1422_3kreps.csv"
[…]

### 2025-09-18 docs/thesis_manuscript/manuscript.tex
cleaning up...

@@
 \title{
-                        \HRule{0.5pt} \\                                                % Upper rule
+%                        \HRule{0.5pt} \\                                                % Upper rule
                         \LARGE \textbf{\uppercase{A Computationally Tractable Model for the Evolution of a Genome's Size and its Growing Fitness Landscape}}      % Title
-                        \HRule{2pt} \\ [0.5cm]          % Lower rule + 0.5cm spacing
+%                        \HRule{2pt} \\ [0.5cm]          % Lower rule + 0.5cm spacing
 %                        \normalsize \today                      % Todays date
@@
 Since life's origins approximately 3.5 billion years ago it has generally increased in complexity, and understanding the processes that generate this complexity is one of the main goals of evolutionary biology \cite{duclosInvestigatingEvolutionDevelopment2019}\cite{mcshea2010biology}.
-One such mechanism of generating complexity is through genome evolution \cite{lynchOriginsGenomeComplexity2003}: by changing the size of their genomes, the earliest living cells - those with a cell membrane, which encode and pass on information through their DNA - were able to create \textit{evolutionary innovations}: new metabolism, physiology or regulation, which augumented their ability to thrive in a variety of environments \cite{barveLatentCapacityEvolutionary2013}\cite{uebbingEvolutionaryInnovationsConserved2024}.
+One such mechanism of generating complexity is through genome evolution \cite{lynchOriginsGenomeComplexity2003}: by changing the size of their genomes, the earliest living cells - those with a cell membrane, which encode and pass on information through their DNA - were able to create \textit{evolutionary innovations}\cite{wagnerMolecularOriginsEvolutionary2011}: new metabolism, physiology or regulation, which augmented their ability to thrive in a variety of environments \cite{barveLatentCapacityEvolutionary2013}\cite{uebbingEvolutionaryInnovationsConserved2024}.
 At the increased cost of replicating and proofreading more DNA, organisms have since been able to evolve new genes allowing them to metabolize different carbon, nitrogen, or complex nutrient sources, various types of receptors and secretion systems to interact, competitively or mutualistically, with others, and the internal regulatory networks required for them to decide which behaviors are optimal under which circumstances, or even to coordinate multicellularity.
@@
 Unlike with the point mutations that are typically observed in every generation, genome evolution occurs up to several orders of magnitude less frequently \cite{sousaRatesTranspositionEscherichia2013}\cite{lipinskiHighSpontaneousRate2011}, and is more heterogenous in its manifestation.
-We define here genome evolution as a mutational process resulting in either the expansion or reduction in the size of the functional genome, arising from one of several causes: amongst genome-increasing mechanisms, there exist duplications from repeat region replication errors \cite{reamsMechanismsGeneDuplication2015}, transposon activity, or whole genome duplications (WGD), as well as genes arising from noncoding sequence through de novo gene birth \cite{sacconeOriginEvolutionGenes2025}, while genome-reducing mechanisms include pseudogenization and deletions via various types of replication errors \cite{burssedMechanismsStructuralChromosomal2022}.
+We define here \textit{genome evolution} as a mutational process resulting in either the expansion or reduction in the size of the functional genome, arising from one of several causes: amongst genome-increasing mechanisms, there exist duplications from repeat region replication errors \cite{reamsMechanismsGeneDuplication2015}, transposon activity, or whole genome duplications (WGD), as well as genes arising from noncoding sequence through de novo gene birth \cite{sacconeOriginEvolutionGenes2025}, while genome-reducing mechanisms include pseudogenization and deletions via various types of replication errors \cite{burssedMechanismsStructuralChromosomal2022}.
 Evolution has been observed to utilize both genome reduction \cite{helsenGeneLossPredictably2020}\cite{szameczGenomicLandscapeCompensatory2014} and expansion \cite{vossebergTimingOriginEukaryotic2021} in order to spur adaptation to increase the organism's \textit{fitness}, conceptualized as some metric representing success e.g.
@@
 
-The fitness landscape was first introduced in 1932 by Sewall Wright to illustrate the vastness of combinitoric space for alleles in a species \cite{wrightRolesMutationInbreeding}.
+The fitness landscape was first introduced in 1932 by Sewall Wright to illustrate the vastness of combinatorial space for alleles in a species \cite{wrightRolesMutationInbreeding}.
 With each unique combination of alleles there can be assigned a fitness value, and so we can generate a topology from a network of possible decision combinations on which we can simulate populations evolving.
@@
 This analogy has spurred a great deal of research in evolutionary biology, where it has found use in modeling macro-level adaptation \cite{venkataramMutualismenhancingMutationsDominate2023}\cite{frankeEvolutionaryAccessibilityMutational2011}, or more recently, in empirical fitness landscapes of proteins \cite{papkouRuggedEasilyNavigable2023}.
-The model can in fact be generalized to represent any sort of decision landscape, and has consequently found itself applied broadly in variety of fields such as directed evolution \cite{alpayEffectsSelectionStringency2024}, theoretical economics \cite{khraishaComplexEconomicProblems2020}, or management studies \cite{siggelkowFirmsSystemsInterdependent2011}.
+The model can in fact be generalized to represent any sort of decision landscape, and has consequently found itself applied broadly in a variety of fields such as directed evolution \cite{alpayEffectsSelectionStringency2024}, theoretical economics \cite{khraishaComplexEconomicProblems2020}, or management studies \cite{siggelkowFirmsSystemsInterdependent2011}.
 Although fitness landscape theory has been used extensively to model adaptation on static genomes or within changing environments, it is less well developed on the scale of long term evolution, where changes in genomic context also arise and can be selected for.
-With this work we extend the Rough Mount Fuji model to incorporate an additional mechanism of genome evolution alongside allelic mutation to explore the effects of multiple mutational processes on evolutionary outcomes.
+With this work we extend the Rough Mount Fuji model to incorporate an additional mechanism of genome evolution, where the dimensionality of the landscape itself can evolve.
+We show that in this model, allowing genome evolution leads to a paradox where maximizing fitness in the short term leads to reduced evolvability in the long term.
 
@@
 The Rough Mount Fuji (RMF) model allows for tunable epistasis \cite{neidhartAdaptationTunablyRugged2014}.
-In the standard model, we represent genotypes as binary vectors, and their permutations in genotype space construct a network where each has a hamming distance of 1 from its neighbors, with the degree of each node equivalent to the number of loci $L$ in the model (see Figure \ref{genotype_space}).
+In the standard model, we represent genotypes as binary vectors, and their permutations in genotype space construct a network where each has a Hamming distance of 1 from its neighbors, with the degree of each node equivalent to the number of loci $L$ in the model (see Figure \ref{genotype_space}).
 
@@
 Genotypes are represented as binary strings e.g.
-[0000], and evolution generally moves over the landscape from low (blue) to high (green) fitness.}
+[0000], and adapting populations generally traverse the landscape from low (blue) to high (green) fitness.}
 	\label{genotype_space}
@@
 
-The assignment of fitness values to genotypes entails sampling two probabilistically determined components: Locus-wise additive effects $A = (a_1, \cdots, a_i) \overset{iid}{\sim} \mathcal{N}(0, \sigma_a)$ are drawn from a normal distribution, and each genotype $g$ recieves a unique epistatic component $e_g \sim \mathcal{N}(0, \sigma_e)$.
+The assignment of fitness values to genotypes entails sampling two probabilistically determined components: Locus-wise additive effects $A = (a_1, \cdots, a_i) \overset{iid}{\sim} \mathcal{N}(0, \sigma_a)$ are drawn from a normal distribution, and each genotype $g$ receives a unique epistatic component $e_g \sim \mathcal{N}(0, \sigma_e)$.
 Fitness values for some arbitrary genotype can then be assigned as the sum of the product of the additive effects and the genotype, plus the epistatic component, as an exponential:
@@
 We model a static $N$ total individuals mutating and reproducing simultaneously in each generation, where each individual may be part of one of many subpopulations of genotypes $g$.
-To initialize the simulation, we specify the maximum number of loci in the model and consequently the length of the vector of additive effects to generate, the simulation length, the variance $\sigma_e$ of the epistatic component, and the mutation rate $\mu$ - the probability for an individual to generate a mutant of hamming distance 1 from its parent.
+To initialize the simulation, we specify the maximum number of loci in the model and consequently the length of the vector of additive effects to generate, the simulation length, the variance $\sigma_e$ of the epistatic component, and the mutation rate $\mu$ - the probability for an individual to generate a mutant of Hamming distance 1 from its parent.
 We proceed to instantiate subsequent generations by first generating mutants from each extant genotype with population $P$ through binomial sampling: $\text{Binomial}(n = P, p = \mu)$.
@@
 
-To allow for genome size changes, we extend the model to give individuals a genome $G$ and a genotype $g$ on that genome, and calculate fitness as a function of both.
-Each $G$ and $g$ are binary strings of the maximum possible length for the genome ($L_{\text{max}}$), and we utilise $G$ to determine which loci in $g$ may mutate: a 1 value in the genome represents the presence of a gene or functional genetic element, and allows the corresponding position in $g$ to be mutated between 0 and 1, while a 0 in $G$ ensures that position in $g$ will always be 0, and thus never contribute additively to fitness.
+The model as described above is static, and assumes all individuals have the same genome.
+To allow for selection to additionally act upon genome size changes, we extend the model to give individuals a genome $G$ and a genotype $g$ on that genome, and calculate fitness as a function of both.
+Each $G$ and $g$ are binary strings of the maximum possible length for the genome ($L_{\text{max}}$), and we utilize $G$ to determine which loci in $g$ may mutate: a 1 value in the genome represents the presence of a gene or functional genetic element, and allows the corresponding position in $g$ to be mutated between 0 and 1, while a 0 in $G$ ensures that position in $g$ will always be 0, and thus never contribute additively to fitness.
 Both mutational processes (genome and genotype mutation) are assigned their own mutation rate parameter ($M$ and $\mu$, respectively), and therefore to access the highest fitness states, populations must first select for genomic changes in $G$ that allow them to mutate their genotypes $g$ to those fitness peaks.
@@
 In this way, rerunning a simulation with identical parameters but varying only the mutation seed allows us to simulate replicates on the same landscape and starting position, that may generate different mutants and thus different outcomes over the course of the simulation.
-Examinations of different classes of landscapes was then performed in separate scripts, each by defining different parameters for the model, and for each replicate, initializing the PRNGs with the appropriate seeds, and running the simulation function.
+Examinations of different classes of landscapes were then performed in separate scripts, each by defining different parameters for the model, and for each replicate, initializing the PRNGs with the appropriate seeds, and running the simulation function.
 
@@
 
-As there is no epistasis, populations will rapidly reach equilibrium at the fitness maximum on any individual landscape, whereas to find the global peak between all lanscapes a slighly different behavior is observed.
+As there is no epistasis, populations will rapidly reach equilibrium at the fitness maximum on any individual landscape, whereas to find the global peak between all landscapes a slightly different behavior is observed.
 Since a mutation to $0$ in either the genotype or genome produces an identical fitness, on average $50\%$ of negative additive effects will be removed via each type of mutation, resulting in half of them being present on the genome with $G_i = 1, g_i = 0$, and half not present with $G_i = g_i = 0$.
@@
 
-As fitnesses are determined solely through the sampling of the epistatic component distribution $\epsilon_{G,g} \overset{iid}{\sim} N(0,\sigma_e)$, larger landscapes will tend to have higher fitness peaks overall, due to a greater number of samples producing, on average, more extreme values.
+As fitnesses are determined solely through the sampling of the epistatic component distribution $e_{G,g}$, larger landscapes will tend to have higher fitness peaks overall, due to a greater number of samples producing, on average, more extreme values.
 This effect can be shown analytically in the expectation of the distribution of fitness peaks, as determined through extreme value theory:  $$\mathbb{E}[X] = \exp{\left(\sqrt{2L \ast (\frac{L \ast \sigma_a^2}{2} + \sigma_e^2) \ast \ln(2)}\right)}$$ Given a constant variance of additive and epistatic terms between landscapes, the mean value of fitness peaks increases with the number of loci as an exponent of $e$, either sub- or superlinearly depending on parameters (Supplementary figure \ref{supp:peak_scaling}).
@@
 Here we see the distributions of fitness values at each genome size split between those replicates which ended the simulation at that size, versus those that continued to expand their genome.
-Consistently among all genome sizes, we notice that replicates that increase their genome size past some size actually have, on average, lower fitnesses that those that do not.
+Consistently among all genome sizes, we notice that replicates that increase their genome size past some size actually have, on average, lower fitnesses than those that do not.
 This implies that populations that begin less fit on some landscape and who minimize their fitness gain on each subsequent landscape possess a greater ability to expand their genome.
+When comparing with figure \ref{ie:sub4}, which displays the same but for uncorrelated landscapes, the magnitude of difference is greater but the fitness values are lower.
+This is in alignment with expected effect of epistasis, which will create more, lower fitness peaks on a landscape, such that it is more likely that populations will become stuck at them.
+Finally, we can confirm in figure \ref{ie:sub3} that the times in which genome size transitions are made is quite similar between those at endpoint at some genome size, and those that continue to expand, suggesting, at least, that genome size transitions are not happening \textit{before} a local maximum is reached on a landscape.
 
@@
         \centering                                                                                                                                                  \subfloat[\label{ie:sub1} Fitness trajectories]{\includegraphics[width = 0.4\textwidth]{figures/genome_trajectories.pdf}}
-        \subfloat[\label{ie:sub2} Endpoints]{\includegraphics[width = 0.4\textwidth]{figures/endpoint_boxplots.pdf}} \\                                             \subfloat[\label{ie:sub3} Size transition timings]{\includegraphics[width = 0.4\textwidth]{figures/timings.pdf}}                                            \subfloat[\label{ie:sub4} Endpoint comparisons on uncorrelated landscapes.]{\includegraphics[width = 0.4\textwidth]{figures/hoc_boxplots.pdf}}              \caption{\textit{(a)}: Fitness trajectories over different genome sizes, \textit{(b)}: fitness comparisons of endpoint and non-endpoint simulations at various genome sizes, \textit{(c)}: Distributions of generation times at which genome size transitions are made, \textit{(d)}: fitness comparisons as in \textit{(b)} on uncorrelated landscapes}
+        \subfloat[\label{ie:sub2} Endpoints]{\includegraphics[width = 0.4\textwidth]{figures/endpoint_boxplots.pdf}} \\                                             \subfloat[\label{ie:sub3} Size transition timings]{\includegraphics[width = 0.4\textwidth]{figures/timings.pdf}}                                            \subfloat[\label{ie:sub4} Endpoint comparisons on uncorrelated landscapes.]{\includegraphics[width = 0.4\textwidth]{figures/hoc_boxplots.pdf}}              \caption{\textit{(a)}: Fitness trajectories over different genome sizes, for replicates with monotonically increasing genome size \textit{(b)}: fitness comparisons of endpoint and non-endpoint simulations at various genome sizes, \textit{(c)}: Distributions of generation times at which genome size transitions are made, \textit{(d)}: fitness comparisons as in \textit{(b)} on uncorrelated landscapes}
         \label{intermediate_epistasis}
@@
 
-	[The results of out simulations reveal some of the new dynamics we observe on expanding landscapes.
-We see how epistasis not only constrains navidability on individual landscapes but the transition between them as well.
-With additive landscapes, there is nothing preventing the collection of all possible additive effects regardless of starting point, whereas when epistasis increases, we see an increased amount of correlation between the starting and ending genome size.
-Most interestingly, we see what happens regarding adaptation over multiple landscapes, who gets the highest fitnesses?
-It's actually the ones that start from lower fitness and minimize fitness gain between when increaseing size that end up with higher fitness.
-We think this is due to an interaction between the scaling of the fitness landscape peak distribution and [xyz].
-As you get fitter, this decreases the probability that any of your neighborw will yield a fitness increasing path, and this applies when jumping between landscapes as well.
-There fore adapting well (to a higher fitness peak) decreases your ability to increase genome size (there is some formula for the uncorrelated landscacpe that shows this).
-However, we also know from examining final fitnesses achieved on larger landscapes, that the mean of the distribution increases, because there are more states in genotype space and therefore more extreme fitnesses have the potential to be generated (This is also sensible.
-The larger the solution space, ther more liekely vert goodsolutions will be found).
+The results of our simulations reveal some of the new dynamics we observe on expanding landscapes.
+We see how epistasis not only constrains navigability on individual landscapes, but the ability to transition between them as well.
+With additive landscapes, populations easily accumulate all possible additive effects regardless of starting genome size, whereas when epistasis increases, we see an increased amount of correlation between the starting and ending sizes.
+Most interestingly, we see what happens regarding adaptation over multiple landscapes with some amount of epistasis: populations that start from lower fitness and minimize fitness gain when increasing genome size end up with overall higher fitnesses.
+We believe this is due to an interaction between the scaling of the fitness landscape peak distribution and adaptability as fitness increases.
+As an individual gets fitter, the probability that any mutant will yield a fitness increasing path decreases, and this also extends to transitions between landscapes as well.
+Therefore adapting to a higher fitness peak on any given landscape actually decreases the probability to be able to increase genome size.
+However, we also know from examining final fitnesses achieved on larger landscapes, that the mean of the distribution increases because there are more states in genotype space, and therefore more extreme fitnesses have the potential to be generated.
 We can then state these two things succinctly:
@@
 These contradictory objectives in the context of epistasis result in the behavior that we see.
-Assuming a low enough genome mutation rate, fitenss peaks will be reached in the current landscape before landscapes will change.
-Thus we can claim that those that hit lower fitness peaks on any landscape, have a greater ability to increase their genome size, and therefore have a higher end probability to have among higher fitnesses, while never having a particularly high fitness (relative to other paths) over the course of the system's evolution.
-]
+Assuming a low enough genome mutation rate, fitness peaks will be reached in the current landscape before landscapes will change.
+Thus, populations that reach lower fitness peaks on a landscape, have a greater ability to increase their genome size, and therefore have a higher end probability to eventually achieve higher fitnesses, while never having a particularly high fitness (relative to other paths) over the course of the system's evolution.
 
@@
 This is very intriguing.
-It suggests that systems that optimize and encode heritable information can actually benefit from avoiding high fitness states too early in the evolution of the system, and finds itself in concordance with results from a similar long-term evolution experiment.
-\cite{woodsSecondOrderSelectionEvolvability2011} To illustrate a biologically relevant scenario in which this would apply, we here introduce a toy model for the evolution of a metabolic network.
+It suggests that systems that optimize and encode heritable information can actually benefit from avoiding high fitness states too early in the evolution of the system, and finds itself in concordance with results from a similar long-term evolution experiment \cite{woodsSecondOrderSelectionEvolvability2011}.
+To illustrate a biologically relevant process where our findings would apply, we here introduce a toy model for the evolution of a metabolic network.
 
@@
 
-[What is the significance of this work?
-Here we show a new model that incorporates an infrequently modeled part of evolution, that the genome can fluctuate both in content and size, and that the behavior seen in this expanded model is not a simple recapitulation of a simpler fitness landscape.
-This matters because I want to demonstrate that due to the darwinian nature of evolution, those lower fitness paths are less likely to be taken in favor of higher fitnesses - this is the consequence of competition.
-Therefore, I would hypothesize that a non-negligible amount of the particular forms life takes now can be attributed to this process.
-I hope that these findings and the model itself allow researchers to  ]
+In this work we show a new RMF model that incorporates a less frequently modeled part of evolution: that the genome can fluctuate in not only allele content but also in size, and that the behavior seen in this expanded model is not a simple recapitulation of a simpler fitness landscape.
+The dynamics observed demonstrate that lower-fitness but more evolvable genomes can achieve the highest fitnesses through genome evolution.
+Despite longer-term adaptational potential, a high strength of selection, especially in larger populations, make these trajectories less likely to survive in a Darwinian environment.
+Therefore, we hypothesize that a non-negligible amount of the particular forms life takes now can be attributed to these sort of contingencies: that highly successful adaptation in earlier environments has, in some cases, limited subsequent evolutionary innovation.
+We hope that these findings, and the model itself, allow researchers to generate more informed hypotheses about long-term genome evolution and help connect empirical findings to create a greater, more cohesive understanding of the interaction between fitness, genome evolution, and adaptation.
 
@@
 
-Special thanks to Suman G. Das for his analytical solution of the fitness peak distribution mean amd the rest of the THEE lab for their valueable comments and critiques throughout this work.
+Special thanks to Suman G. Das for his analytical solution of the fitness peak distribution mean and the rest of the THEE lab for their valuable comments and critiques throughout this work.
 
